Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Data

### 2017-09-18 article.Rnw
Spell check

\author{
  Eduardo E. R. Junior\thanks{Corresponding author:
    \href{mailto:jreduardo@usp.br}{\nolinkurl{jreduardo@usp.br}}}
    \hspace*{.8mm}\thanks{Department of Exact Sciences, ESALQ-USP,
    Piracicaba, São Paulo, Brazil} \and
  Walmes M. Zeviani\thanks{Department of Statistics, UFPR, Curitiba,
    Paraná, Brazil} \and
  Wagner H. Bonat\footnotemark[3] \and
  Clarice G. B. Dem\'{e}trio\footnotemark[2]
}

@@

  In the analysis of count data seldom the equidispersion assumption is
  suitable, hence the Poisson regression model is inappropriate in many
  situations. The COM-Poisson is a member of the exponential family, has
  the Poisson and geometric distributions as special cases and the
  Bernoulli distribution as a limiting case. It can deal with both under
  and overdispersed count data. The main disadvantage of COM-Poisson
  distribution is that the location parameter does not represent the
  average making difficult interpretation in a regression model. In this
  paper we present a very simple reparametrization based on mean
  approximation. The estimation and inference is based on likelihood and
  simulation study showed that the maximum likelihood estimators are
  unbiased and consistent for both regression and precision
  parameters. Furthermore we observe empirical correlation between
  regression and precision parameters close to 0 suggest the
  orthogonality for the proposed reparametrization. We presents
  applications of the proposed model by three case studies with over,
  under and equidispersed counts. Results show improvements over the
  conventional approaches. Therefore, its use in the analysis of count
  data is encouraged. The computational routines for fitting of original
  and reparametrized COM-Poisson models are available in the
  supplementary material.

@@

In practice, overdispersion, when the variance greater than the mean, is
largely reported in the literature and may occur due to the absence of
relevant covariates, heterogeneity of sampling units, different domain
not considered and excess of zero \citep{Hinde1998}. Lesser reported,
but that has received attention from the statistical community, is a
underdispersion case, when the variance is less than the mean. The
process that reduce the variability are not as well known that lead to
extra variability. For the same reason, there a few approaches to deal
underdispersed cases. The explanatory mechanisms to underdispersion may
be related to Poisson process. When the time between events is not
exponential distributed, the number of events can be over or
underdispersed, this motivated the class of duration dependence models
\citep{Winkelmann1995}. Other possibity to explain underdispersion are
order statistics. Suppose $Y_i$ is Poisson distributed then maximum of
$Y_i$ is underdispersed.

@@

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and \cite{Huang2017} parametrization. The
proposed reparametrization and assessment of moments approximations is
considered in the \autoref{reparametrization}. In the
\autoref{estimation-and-inference} we show the estimation and inference
methods based on likelihood function whose properties are assess in
\autoref{simulation-study} by simulations. Applications of
reparametrized COM-Poisson regression model are illustrated in
\autoref{case-studies} with three case studies. The analysis are
performed using the software R \citep{Rcore2017}. The source code and
data sets are avaliable in supplementary material\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

In this section we conducted a simulation study to assess the properties
of the proposed model and estimators. We considered average counts
varying from 3 to 27 according to a regression model structure with a
continuous variable $\bm{x}_1$ between 0 and 1 and a categorical
variable $\bm{x}_2$ with three levels. Then the systematic componente is
given by $\bm{\mu} = \exp(\beta_0 + \beta_1 \bm{x}_1 + \beta_{21}
\bm{x}_{21} + \beta_{22} \bm{x}_{22})$, where $\bm{x}_{21}$ and
$\bm{x}_{22}$ are \textit{dummy} variable for $\bm{x}_2$. With respect
to the precision parameter $\phi$, we fixed it to four values $-1.6$,
$-1.0$, $0.0$ and $1.8$, in order to have strong overdispersed, moderate
overdispersed, equidispersed and underdispersed counts
respectively. \autoref{fig:justpars} shows the variation of the average
counts (left) and variation of dispersion index (E$(Y)$ / var$(Y)$) for
each fixed value of the precision parameter $\phi$. This configuration
allows us to assess the proprierties of the estimators in extreme
situation such as This configuration allows us to evaluate the
properties of the estimators in extreme situations such as high counts
and low dispersion, and low counts and high dispersion.

@@

In this paper, we argued the parametrization of COM-Poisson distribution
to analyse dispersed count data. We proposed a novel reparametrization
based only mean approximation gives by \cite{Shmueli2005}. The
likelihood-based method to estimation and inference is presented. We
assess the properties of these estimators by a simulation
study. Applications of the proposed model are showed by three case
studies with over, under and equidispersed counts. In the data analysis
we compared standard Poisson, original COM-Poisson, repametrized
COM-Poisson and quasi-Poisson models.

We computed the quadratic errors for the mean approximation. The results
showed that it is well accurate supports our proposed
reparametrization. The results of a simulation study showed that the
estimators of the regression and precision parameters are consistent and
unbiased and asymptotically unbiased respectively. The empirical
coverage rates of the confidence intervals for the parameters, compute
by the asymptotical distrition of the estimators, are close to the
nominal level for sizes greather than 100. The worst scenario is when we
have small sample size and strong overdispersed counts. In general, we
recommend the use of Wald intervals for computational simplicity and
good results.

The data analysis in the case studies has shown that COM-Poisson is a
suitable distribution for over, under and equidispersion situations. The
observed empirical correlation between regression and precision
parameters suggest the orthogonality between $\mu$ and $\phi$ in
COM-Poisson$\mu$ distribution against the original parametrization. Due
to this feature the computational procedure is faster under the
reparameterized model. Another advantage, perhaps the main, of the
reparametrization is that the estimated regression parameters have
interpretation of rate ratios, same of the Poisson model. In comparison
of the quasi-Poisson approach the results was similar in both
simulations, nevertheless the COM-Poisson is a full parametric model,
allowing the calculation of probabilities for example.

In general, the results presented by the reparameterized COM-Poisson
models were satisfactory and superior to the conventional
approaches. Therefore, its use in the analysis of count data is
encouraged. The computational routines for fitting of original and
reparametrized COM-Poisson models are available in the supplementary
material\textsuperscript{\ref{papercompanion}}.

There are many possible extensions and possible topics for futher
investigation to model discussed in the present paper, including
simulation study to assess model robustness against distribution
miss-specification, asses theoretical approximations for
$Z(\lambda, \nu)$ (or $Z(\mu,\phi))$, in order to avoid the selection of
sum's upper bound, propose a double GLM based on the reparametrized
COM-Poisson model and a accomodate random effects since we have a model
parametrized on mean.

>>>>

\author{
  Eduardo E. R. Junior\thanks{Corresponding author:
    \href{mailto:jreduardo@usp.br}{\nolinkurl{jreduardo@usp.br}}}
    \hspace*{.8mm}\thanks{Department of Exact Sciences, ESALQ-USP,
    Piracicaba, S\~{a}o Paulo, Brazil} \and
  Walmes M. Zeviani\thanks{Department of Statistics, UFPR, Curitiba,
    Paran\'{a}, Brazil} \and
  Wagner H. Bonat\footnotemark[3] \and
  Clarice G. B. Dem\'{e}trio\footnotemark[2]
}

@@

  In the analysis of count data seldom the equidispersion assumption is
  suitable, hence the Poisson regression model is inappropriate in many
  situations. The COM-Poisson is a member of the exponential family, has
  the Poisson and geometric distributions as special cases and the
  Bernoulli distribution as a limiting case. It can deal with both under
  and overdispersed count data. Therefore it is a suitable distribution
  to analysis of dispersed count data. In spite of the nice properties
  of COM-Poisson, the location parameter of the distribution does not
  represent the average making difficult interpretation in a regression
  structure. In this paper we present a very simple reparametrization
  based on mean approximation. The estimation and inference is based on
  likelihood and simulation study showed that the maximum likelihood
  estimators are unbiased and consistent for both regression and
  precision parameters. Furthermore, we observe empirical correlation
  between regression and precision parameters close to zero, suggest the
  orthogonality feature in the proposed reparametrization model. We
  presents applications of the proposed model by three case studies with
  over, under and equidispersed counts. Results show improvements over
  the conventional approaches. Therefore, its use in the analysis of
  count data is encouraged. The computational routines for fitting of
  original and reparametrized COM-Poisson models are available in the
  supplementary material.

@@

In practice, overdispersion, when the variance greater than the mean, is
largely reported in the literature and may occur due to the absence of
relevant covariates, heterogeneity of sampling units, different domain
not considered and excess of zero \citep{Hinde1998}. Lesser reported,
but that has received attention from the statistical community, is a
underdispersion case, when the variance is less than the mean. The
process that reduce the variability are not as well known that lead to
extra variability. For the same reason, there a few approaches to deal
underdispersed cases. The explanatory mechanisms to underdispersion may
be related to Poisson process. When the time between events is not
exponential distributed, the number of events can be over or
underdispersed, this motivated the class of duration dependence models
\citep{Winkelmann1995}. Other possibility to explain underdispersion are
order statistics. Suppose $Y_i$ is Poisson distributed then maximum of
$Y_i$ is underdispersed.

@@

This paper is organized as follows. In \autoref{background} we present
the COM-Poisson distribution and \cite{Huang2017} parametrization. The
proposed reparametrization and assessment of moments approximations is
considered in the \autoref{reparametrization}. In the
\autoref{estimation-and-inference} we show the estimation and inference
methods based on likelihood function whose properties are assess in
\autoref{simulation-study} by simulations. Applications of
reparametrized COM-Poisson regression model are illustrated in
\autoref{case-studies} with three case studies. The analysis are
performed using the software R \citep{Rcore2017}. The source code and
data sets are available in supplementary material\footnote{Available on
  \texttt{\url{http://www.leg.ufpr.br/~eduardojr/papercompanions}}
  \label{papercompanion}.}.

@@

In this section we conducted a simulation study to assess the properties
of the proposed model and estimators. We considered average counts
varying from 3 to 27 according to a regression model structure with a
continuous variable $\bm{x}_1$ between 0 and 1 and a categorical
variable $\bm{x}_2$ with three levels. Then the systematic component is
given by $\bm{\mu} = \exp(\beta_0 + \beta_1 \bm{x}_1 + \beta_{21}
\bm{x}_{21} + \beta_{22} \bm{x}_{22})$, where $\bm{x}_{21}$ and
$\bm{x}_{22}$ are \textit{dummy} variable for $\bm{x}_2$. With respect
to the precision parameter $\phi$, we fixed it to four values $-1.6$,
$-1.0$, $0.0$ and $1.8$, in order to have strong overdispersed, moderate
overdispersed, equidispersed and underdispersed counts
respectively. \autoref{fig:justpars} shows the variation of the average
counts (left) and variation of dispersion index (E$(Y)$ / var$(Y)$) for
each fixed value of the precision parameter $\phi$. This configuration
allows us to assess the properties of the estimators in extreme
situation such as This configuration allows us to evaluate the
properties of the estimators in extreme situations such as high counts
and low dispersion, and low counts and high dispersion.

@@

In this paper, we argued the parametrization of COM-Poisson distribution
to analyse dispersed count data. We proposed a novel reparametrization
based only mean approximation gives by \cite{Shmueli2005}. The
likelihood-based method to estimation and inference is presented. We
assess the properties of these estimators by a simulation
study. Applications of the proposed model are showed by three case
studies with over, under and equidispersed counts. In the data analysis
we compared standard Poisson, original COM-Poisson, reparametrized
COM-Poisson and quasi-Poisson models.

We computed the quadratic errors for the mean approximation. The results
showed that it is well accurate supports our proposed
reparametrization. The results of a simulation study showed that the
estimators of the regression and precision parameters are consistent and
unbiased and asymptotically unbiased respectively. The empirical
coverage rates of the confidence intervals for the parameters, compute
by the asymptotic distribution of the estimators, are close to the
nominal level for sizes greater than 100. The worst scenario is when we
have small sample size and strong overdispersed counts. In general, we
recommend the use of Wald intervals for computational simplicity and
good results.

The data analysis in the case studies has shown that COM-Poisson is a
suitable distribution for over, under and equidispersion situations. The
observed empirical correlation between regression and precision
parameters suggest the orthogonality between $\mu$ and $\phi$ in
COM-Poisson$\mu$ distribution against the original parametrization. Due
to this feature the computational procedure is faster under the
reparametrized model. Another advantage, perhaps the main, of the
reparametrization is that the estimated regression parameters have
interpretation of rate ratios, same of the Poisson model. In comparison
of the quasi-Poisson approach the results was similar in both
simulations, nevertheless the COM-Poisson is a full parametric model,
allowing the calculation of probabilities for example.

In general, the results presented by the reparametrized COM-Poisson
models were satisfactory and superior to the conventional
approaches. Therefore, its use in the analysis of count data is
encouraged. The computational routines for fitting of original and
reparametrized COM-Poisson models are available in the supplementary
material\textsuperscript{\ref{papercompanion}}.

There are many possible extensions and possible topics for further
investigation to model discussed in the present paper, including
simulation study to assess model robustness against distribution
miss-specification, asses theoretical approximations for
$Z(\lambda, \nu)$ (or $Z(\mu,\phi))$, in order to avoid the selection of
sum's upper bound, propose a double GLM based on the reparametrized
COM-Poisson model and a accommodate random effects since we have a model
parametrized on mean.
